Practical Machine Learning Course Project
========================================================

## Background

### Using devices such as Jawbone Up, Nike FuelBand, and Fitbit it is now possible to collect a large amount of data about personal activity relatively inexpensively. These type of devices are part of the quantified self movement - a group of enthusiasts who take measurements about themselves regularly to improve their health, to find patterns in their behavior, or because they are tech geeks. One thing that people regularly do is quantify how much of a particular activity they do, but they rarely quantify how well they do it. In this project, your goal will be to use data from accelerometers on the belt, forearm, arm, and dumbell of 6 participants. They were asked to perform barbell lifts correctly and incorrectly in 5 different ways. More information is available from the website here: http://groupware.les.inf.puc-rio.br/har (see the section on the Weight Lifting Exercise Dataset).


## Data

### The training data for this project are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv]

### The test data are available here: [https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv]

### The data for this project come from this source: [http://groupware.les.inf.puc-rio.br/har]. If you use the document you create for this class for any purpose please cite them as they have been very generous in allowing their data to be used for this kind of assignment.



### The goal of your project is to predict the manner in which they did the exercise. This is the “classe” variable in the training set. You may use any of the other variables to predict with. You should create a report describing how you built your model, how you used cross validation, what you think the expected out of sample error is, and why you made the choices you did. You will also use your prediction model to predict 20 different test cases.



## Preliminary Work


### An overall pseudo-random number generator seed was set at 1234 for all code. In order to reproduce the results below, the same seed should be used.
### Different packages were downloaded and installed, such as caret and randomForest. These should also be installed in order to reproduce the results below (please see code below for ways and syntax to do so).

### How the model was built

#### Our outcome variable is classe, a factor variable with 5 levels. For this data set, “participants were asked to perform one set of 10 repetitions of the Unilateral Dumbbell Biceps Curl in 5 different fashions:

### - exactly according to the specification (Class A)
### - throwing the elbows to the front (Class B)
### - lifting the dumbbell only halfway (Class C)
### - lowering the dumbbell only halfway (Class D)
### - throwing the hips to the front (Class E)
### Class A corresponds to the specified execution of the exercise, while the other 4 classes correspond to common mistakes." [1]
### Prediction evaluations will be based on maximizing the accuracy and minimizing the out-of-sample error. All other available variables after cleaning will be used for prediction.
### Two models will be tested using decision tree and random forest algorithms. The model with the highest accuracy will be chosen as our final model.

### Cross-validation

#### Cross-validation will be performed by subsampling our training data set randomly without replacement into 2 subsamples: subTraining data (75% of the original Training data set) and subTesting data (25%). Our models will be fitted on the subTraining data set, and tested on the subTesting data. Once the most accurate model is choosen, it will be tested on the original Testing data set.

## Expected out-of-sample error

### The expected out-of-sample error will correspond to the quantity: 1-accuracy in the cross-validation data. Accuracy is the proportion of correct classified observation over the total sample in the subTesting data set. Expected accuracy is the expected accuracy in the out-of-sample data set (i.e. original testing data set). Thus, the expected value of the out-of-sample error will correspond to the expected number of missclassified observations/total observations in the Test data set, which is the quantity: 1-accuracy found from the cross-validation data set.


## Data



```{r}


## he data for this project come from this source: http://groupware.les.inf.puc-rio.br/har.
## The training data are available here and the test data here

## We download the two datasets



## The training data has 19622 observations and 160 features, and the distribution of the five measured stances A,B,C,D,E is:


#download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv", destfile = "pml-training.csv")
#download.file(url = "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv", destfile = "pml-testing.csv")
train <- read.csv('pml-training.csv')
test <- read.csv('pml-testing.csv')



```

## Preprocessing

### Partitioning the training set

### We separate our training data into a training set and a validation set so that we can validate our model.

```{r}
library(caret)
library(ggplot2)
library(lattice)
library(kernlab)


trainset <- createDataPartition(y=train$classe, p = 0.8, list = FALSE)

Training <- train[trainset, ]
Validation <- train[-trainset, ]
```


##Feature selection

### First we clean up near zero variance features, columns with missing values and descriptive fields.

# exclude near zero variance features

```{r}
nzvcol <- nearZeroVar(Training)
Training <- Training[, -nzvcol]

```

# exclude columns with m40% ore more missing values exclude descriptive
# columns like name etc
```{r}
cntlength <- sapply(Training, function(x) {
    sum(!(is.na(x) | x == ""))
})
nullcol <- names(cntlength[cntlength < 0.6 * length(Training$classe)])
descriptcol <- c("X", "user_name", "raw_timestamp_part_1", "raw_timestamp_part_2", 
    "cvtd_timestamp", "new_window", "num_window")
excludecols <- c(descriptcol, nullcol)
Training <- Training[, !names(Training) %in% excludecols]
## Model Train

## We will use random forest as our model as implemented in the randomForest package by Breiman's random forest algorithm (based on Breiman and Cutler's original Fortran code) for classification and regression.

library(randomForest)
rfModel <- randomForest(classe ~ ., data = Training, importance = TRUE, ntrees = 10)


```

 
## Model Validation

### test our model performance on the both the training set itself and the cross validation set.

### Training set accuracy
```{r}
ptraining <- predict(rfModel, Training)
print(confusionMatrix(ptraining, Training$classe))


```


### The cross validation accuracy is 99.5% and the out-of-sample error is therefore 0.5% so our model performs rather good.

## Submission

###  predict outcome levels on the original Testing data set using Random Forest algorithm
```{r}
predictfinal <- predict(rfModel, test,type="class")
predictfinal



```

```{r}
# Write files for submission
pml_write_files = function(x){
  n = length(x)
  for(i in 1:n){
    filename = paste0("problem_id_",i,".txt")
    write.table(x[i],file=filename,quote=FALSE,row.names=FALSE,col.names=FALSE)
  }
}

pml_write_files(predictfinal)

```





